What is an Ancestral Recombination Graph?

### 2022-02-11 paper.tex
working through ARG representation section

@@

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format.
The concrete details of how an ARG should be stored and processed
are not discussed in the literature, % weak, fix.
but the following is a reasonable approach that mirrors the implementations
in methods such as [x, y z].
An ARG is composed of a set of $N$ nodes, where each node is an
integer $0 \leq u < N$, and edges $(u, v)$ define the relatships between
those nodes.
Each node corresponds to an \emph{event}, and is either a common ancestor
event (in which case it has two children and one parent)
or a recombination event (when it has one child and two parents).
Each recombination event is associated with a breakpoint $0 < x < L$,
which we might encode as a mapping $f: N \rightarrow \{0, ..., L - 1\}$
such that $f(u)$ is the recombination breakpoint if $u$ is a recombination
node and $f(u) = 0$ otherwise.

>>>>

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format.
The concrete details of how an ARG should be stored and processed
are not discussed in the literature, % weak, fix.
but the following is a reasonable approach that mirrors the implementations
in methods such as [x, y z].
A Griffiths ARG is composed of a set of $N$ nodes, where each node is an
integer $0 \leq u < N$, and edges $(u, v)$ define the relationships between
those nodes (Fig.~\ref{fig-arg-data-structure}A).
Each node corresponds to an \emph{event}, and is either a common ancestor
event (in which case it has two children and one parent)
or a recombination event (when it has one child and two parents).
Each recombination event is associated with a breakpoint $0 < x < L$,
which we might encode as a mapping $f: N \rightarrow \{0, ..., L - 1\}$
such that $f(u)$ is the recombination breakpoint if $u$ is a recombination
node and $f(u) = \infty$ otherwise. This encoding captures the information
of a realisation of the ARG stochastic process, but it is important to
note that it does not uniquely describe the underlying ancestral history
without stipulating some additional ordering requirements.

@@

The Griffiths ARG encoding is ambiguous because it does not explicitly
define which of the parent lineages resulting from a recombination
event carries ancestral material to the left or right of the
breakpoint. The ambiguity is resolved by adopting a simple rule when
traversing the graph in order to generate the marginal trees for a
given position $z$: at a recombination node $u$, we follow the
first edge in which $u$ is a child if $f(u) \leq z$, and the
second edge otherwise. This ordering requirement, which straightforward
% Saying "awkward" too many times here, rephrase.
to describe can be awkward in practise. For example,
simulating an ARG in this representation is considerably
complicated by the fact that the
first event to occur may hit the right-hand lineage rather
than the left.
Such awkwardnesses can be worked around, of course,
but in practise methods such has
\texttt{beagle}~\citep{lyngso2005minimum} and
\texttt{KwARG}~\citep{ignatieva2021kwarg}
resolve this problem by
associating a label ``P'' (prefix) or ``S'' (suffix) with the
outbound edges of a recombination event.

% it's not reasonable to assume that one type of event happens
% at a time any more.
The Griffiths ARG encoding is also limited in the types of ancestral
history that can be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$ and [some assumptions about genome length]
\textcolor{red}{what is this Jere?}. While these approximations are
reasonable
to make when studying the Kreitman \emph{Drosphila Melanogaster}
dataset~\citep{kreitman1983nucleotide} ($n=11$, $N\approx 10^6$,
at 43 variant sites over x Kb \textcolor{red}{details and citation
for Ne estimate}) that is classically used for benchmarking
ARG inference methods [cites] , they are highly questionable in
modern datasets. For example, there are now many
human datasets consisting of hundreds of thousands
of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
While the coalescent model is suprisingly robust to such
basic violations to the underlying assumptions~\citep{wakeley2012gene},
the combination of a large sample size and long sequence length
can produce pathological behaviour in
simulations~\citep{bhaskar2014distortion,nelson2020accounting}.
